---
title: "Dispersal statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dispersal_statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  fig.width = 7.5,
  fig.height = 4.5
)
```

```{r, message = FALSE}
# load the package into R
library(phyloSpreadR)

# load other useful packages
library(ggtree)
library(ggplot2)
library(toolkitSeqTree)
library(RColorBrewer)
```

First we read a tree into R and "fortify" using `ggtree::fortify()` the tree object which produces a data frame with one row representing each tree node.

```{r}
beast <- treeio::read.beast('../inst/extdata/study2.mcc.trees')
beast_dat <- fortify(beast)
```

This script reads in maximum clade credibility tree from continuous BEAST analysis carried out in beast. Each node of the phylogeny is labelled with: `branch.length`, the time duration of branch in years; `height`, the time in years before the most recent sampling date (MRSD); `location1`, the sampled or estimated latitude; and `location2`, the sampled or estimated longitude.

### Process spatio-temporal information

Dates for each node are calculated using `height` and the most recent sampling date (MRSD).

```{r}
MRSD <- "2017-12-11"
beast_dat$date <- Date2decimal(MRSD) - beast_dat$height
```

Distances are calculated from geographic coordinates.

```{r}
beast_dat <- process_spatiotemporal(beast_dat)
root_node <- beast_dat$node[beast_dat$node == beast_dat$parent]
```

#### Distances

Distance covered per branch can now be plotted for each branch/node.

```{r}
## Summary of distribution of per-branch distances
summary(beast_dat$gcd)
ggplot(beast_dat[beast_dat$node != root_node,], aes(x = gcd)) +
  geom_histogram(bins = 50) +
  labs(x = "Distance (km)", y = "Frequency") +
  theme_minimal()

## Colour branches on tree by distance covered
ggtree(beast, mrsd = MRSD, aes(col = beast_dat$gcd)) +
  scale_colour_gradientn(colours = brewer.pal(11, "Spectral"),
                         name = 'Distance\n(km)') +
  theme_tree2()
```

#### Distance to epidemic source

The function `process_spatiotemporal()` also calculates the distance from the estimated location of the root for the phylogeny for each other node in the tree.

```{r}
summary(beast_dat$dist_root)
ggplot(beast_dat, aes(x = dist_root)) +
  geom_histogram(bins = 30) +
  labs(x = "Distance from root (km)", y = "Frequency") +
  theme_minimal()

## Colour nodes on tree by distance to root
ggtree(beast, mrsd = MRSD, col = 'grey40') +
  geom_point(aes(fill = beast_dat$dist_root), pch = 21, size = 2) +
  scale_fill_gradientn(colours = brewer.pal(11, "Spectral"),
                       name = 'Distance\nfrom\nroot (km)') +
  theme_tree2()
```

#### Epidemic wavefront

For each node, distances from the root and the dates of internal nodes can be used to estimate the maximal spatial wavefront distance from the epidemic origin through time.

```{r}
ggplot(beast_dat, aes(x = date, y = wavefront)) +
  geom_line() +
  labs(x = 'Date', y = 'Wavefront distance (km)') +
  theme_minimal()
```

This graph would appear smoother when based on multiple trees in posterior sample rather than just the maximum clade credibility.

#### Velocity and diffusion

The function `process_spatiotemporal()` also calculates the velocity and diffusion associated with each branch. Velocity is calculated as distance / time for the branch whereas diffusion is calculated as distance^4 / time^2.

```{r}
## velocity on tree
ggtree(beast, mrsd = MRSD, aes(col = beast_dat$velocity / 365)) +
  scale_colour_gradientn(colours = brewer.pal(11, "Spectral"), trans = 'log',
                       name = 'Velocity\n(km/day)') +
  theme_tree2()

## diffusion on tree
ggtree(beast, mrsd = MRSD, aes(col = beast_dat$diffusion / 365)) +
  scale_colour_gradientn(colours = brewer.pal(11, "Spectral"), trans = 'log',
                       name = 'Diffusion\n(km^2/day)') +
  theme_tree2()

```

### Dispersal statistics

Each branch represnts a reconstructed movement with a distance, duration, velocity and diffusion rate. Summary statistics can be calculated using functions

#### Weighted and un-weighted dispersal velocity

```{r}
calc_dispersal_velocity(beast_dat)

calc_dispersal_velocity(beast_dat, weighted = TRUE)
```

#### Weighted and un-weighted diffusion rate

```{r}
calc_diffusion_coef(beast_dat)

calc_diffusion_coef(beast_dat, weighted = TRUE)
```

```{r}
dispersal_stats(beast_dat)
```

### Clade-specific statistics

It may be desirable to calculate statistics for a specific clade in the phylogeny or even another selection of branches selected using an alternative criterion.

Each node of the tree is associated with a node number:

```{r}
ggtree(beast) +
  geom_text(aes(label=node))
```

Clades defined by certain nodes can be subsetted using `toolkitSeqTree::node_descendants`

```{r}
descendants_127 <- node_descendants(beast_dat, node_id = 127)
descendants_93 <- node_descendants(beast_dat, node_id = 93)

ggtree(beast) +
  geom_point(data = subset(beast_dat, node == 127),
             col = 'firebrick2', size = 4) +
  geom_point(data = subset(beast_dat, node %in% descendants_127),
             col = 'firebrick2', shape = 18, size = 2) +
  geom_point(data = subset(beast_dat, node == 93),
             col = 'dodgerblue3', size = 4) +
  geom_point(data = subset(beast_dat, node %in% descendants_93),
             col = 'dodgerblue3', shape = 18, size = 2)
```

The functions used above to calculate dispersal statistics can be supplied with either a vector of nodes or a single target node which will be used to pull out a clade.

```{r}
# by default function returns stat for whole tree
calc_dispersal_velocity(beast_dat, weighted = TRUE)

# value for either of the clades plotted above can be returned by
# or the node defining them
calc_dispersal_velocity(beast_dat, weighted = TRUE, target_node = descendants_127)
calc_dispersal_velocity(beast_dat, weighted = TRUE, target_node = 127)
```

Target node can also be supplied to `dispersal_stats`.

```{r}
dispersal_stats(beast_dat)
dispersal_stats(beast_dat, target_node = 127)
dispersal_stats(beast_dat, target_node = 98)
```


### Clade-specific wavefront

In addition to calculating dispersal summary statistics for a given clade, it might be desirable to calculate the distance from the 'root' of a clade and the epidemic wavefront within the clade.

```{r}
beast_dat <- clade_wavefront(beast_dat, target_node = 127)

## Colour nodes on tree by distance to root
ggtree(beast, mrsd = MRSD, col = 'grey40') +
  geom_point(aes(fill = beast_dat$dist_node_127),
             pch = 21, size = 2) +
  scale_fill_gradientn(colours = brewer.pal(11, "Spectral"),
                       name = 'Distance\nfrom\ntarget\nnode (km)',
                       na.value = 'grey80') +
  theme_tree2()

ggplot(beast_dat[is.na(beast_dat$wavefront_node_127) == F,],
       aes(x = date, y = wavefront_node_127)) +
  geom_line() +
  labs(x = 'Date', y = 'Clade-specific wavefront distance (km)') +
  theme_minimal()
```
